---
title: "Week Four - Incumbency"
author: "Grant Williams"
date: "2024-10-01"
output: pdf_document
categories: []
tags: []
slug: "week-four-incumbency"
---

# Introduction

In this fourth blog post, I want to explore the subject of *incumbency.*

```{r, include = FALSE, warning = FALSE, message = FALSE}

# Load libraries.
## install via `install.packages("name")`
library(car)
library(caret)
library(CVXR)
library(glmnet)
library(kableExtra)
library(maps)
library(RColorBrewer)
library(sf)
library(tidyverse)
library(viridis)
library(stargazer)

####----------------------------------------------------------#
#### Read, merge, and process data.
####----------------------------------------------------------#

# Read incumbency/vote data.
d_vote <- read_csv("popvote_1948-2020.csv")
d_state_vote <- read_csv("state_popvote_1948_2020.csv")
d_vote$party[d_vote$party == "democrat"] <- "DEM"
d_vote$party[d_vote$party == "republican"] <- "REP"

# Read economic data.
d_econ <- read_csv("fred_econ.csv") |> 
  filter(quarter == 2)

# Read polling and election results data. 
d_pollav_natl <- read_csv("national_polls_1968-2024.csv")
d_pollav_state <- read_csv("state_polls_1968-2024.csv")

# Shape and merge polling and election data using November polls. 
d_poll_nov <- d_vote |> 
  left_join(d_pollav_natl |> 
              group_by(year, party) |> 
              top_n(1, poll_date) |> 
              select(-candidate), 
            by = c("year", "party")) |> 
  rename(nov_poll = poll_support) |> 
  filter(year <= 2020) |> 
  drop_na()

# Create dataset of polling average by week until the election. 
d_poll_weeks <- d_pollav_natl |> 
  group_by(year, party, weeks_left) |>
  summarize(mean_poll_week = mean(poll_support)) |> 
  filter(weeks_left <= 30) |> 
  pivot_wider(names_from = weeks_left, values_from = mean_poll_week) |> 
  left_join(d_vote, by = c("year", "party"))

####----------------------------------------------------------#
#### Descriptive statistics on the incumbency advantage. 
####----------------------------------------------------------#
# A different way of assessing the incumbency advantage 
# (out of 11 elections where there was at least one incumbent running). 
inc_tab <- d_vote |> 
  filter(year > 1948 & year < 2024) |>
  select(year, party, candidate, incumbent, winner) |> 
  pivot_wider(names_from = party, 
              values_from = c(candidate, incumbent, winner)) |> 
  filter(incumbent_DEM == TRUE | incumbent_REP == TRUE)

# I AM USING CHATGPT TO MAKE THIS PI CHART
# Add a new column indicating if the incumbent won or lost
inc_tab <- inc_tab %>%
  mutate(incumbent_won = ifelse((incumbent_DEM & winner_DEM) | (incumbent_REP & winner_REP), "Won", "Lost"))

# Create a new data frame for pie chart labels
label_data <- inc_tab %>%
  group_by(incumbent_won, year) %>%
  summarise(count = n(), .groups = 'drop')
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

# Create the pie chart with year labels
ggplot(label_data, aes(x = "", y = count, fill = incumbent_won)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_bar(stat = "identity", width = 1, color = "black") + 
  geom_text(aes(label = year), position = position_stack(vjust = 0.5), size = 3, color = "white") +
  labs(title = "Incumbency Advantage since WWII (1948-2020)",
       fill = "Incumbent Outcome") +
  theme_void() +
  scale_fill_manual(values = c("Won" = "darkgreen", "Lost" = "red2"))
```

In the 11 presidential elections since 1948 in which the incumbent president was running, the incumbent's bid was successful 7 times. As a percentage, this is roughly 64% of the time.

Political pundits have long identified this apparent benefit of currently holding the political office for which one is running as *incumbency advantage.*

One need only take a look at the House reelection rates among incumbent candidates to see clear evidence of this advantage.

```{r, echo = FALSE, warning = FALSE, message = FALSE}

house_incumbency <- read_csv("HouseIncumbency.csv")

colnames(house_incumbency) <- c("Year", "Rate")

avg <- mean(house_incumbency$Rate)

ggplot(house_incumbency, aes(x = factor(Year), y = Rate)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(title = "House Incumbency Reelection by Election Cycle (1964 to 2022)",
       x = "Election Cycle",
       y = "Incumbency Election (%)",
       subtitle = "Average Success Rate is 93.16%",
       caption = "Source: OpenSecrets.org") +
  geom_hline(yintercept = avg, color = "red", linetype = "dashed") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))



```

Between 1964 and 2022, on average, incumbent candidates running for the House of Representatives won their election 93.16% of the time ([OpenSecrets](https://www.opensecrets.org/elections-overview/reelection-rates)). While this percentage is not wholly attributable to incumbency advantage since most House elections occur in non-competitive congressional districts, this 93% figure still conveys the political edge that incumbency affords political candidates. 

Many have hypothesized that incumbency advantage is largely the product of incumbent candidates' name-recognition, previous experience campaigning and fundraising, and their ability to take credit for economic gains over the last 4 years. It is this last element of incumbency advantage, also known as *pork*, that I want to focus my blog on. 

The code used to produce these visualizations is publicly available in my [github repository](https://github.com/grantbw4/election-blog) and draws heavily from the section notes and sample code provided by the Gov 1347 Head Teaching Fellow, [Matthew Dardet](https://www.matthewdardet.com/harvard-election-analytics-2024/).

## Analysis

The term, **"pork,"** was first used to reference questionably appropriated public spending in the 19th-century by American author Edward Everett Hale in his 1863 short story, "The Children of the Public." Contained within his larger book. *The Man Without a Country*, "The Children of the Public" uses the imagery of pork barrels to suggest wasteful spending. At a time when refrigerators were not yet invented, pork would be salted and stored in large, wooden barrels. These wooden barrels were quite valuable. Hence, "pork barrel spending" is like taking a cut from this larger pool of resources. Examples of "pork" include earmarked investments in unimportant projects that go toward specific states or constituents, often to gain their favor or reward them for their support. 

The theory of pork in the context of incumbency advantage is that presidential candidates can propose budgets and lead initiatives that will direct federal funding to more competitive states that they want to win in the next election cycle. 

To test this, I will first visualize federal grant spending in millions of dollars by states to see if competitive states receive more money than non-competitive states, on average. I will do this by using a 1984-2008 federal grants data set harmonized by Douglas Kriner and Andrew Reeves as part of their 2015 paper, ["Presidential Particularism and Divide-the-Dollar Politics."](https://www-cambridge-org.ezp-prod1.hul.harvard.edu/core/journals/american-political-science-review/article/presidential-particularism-and-dividethedollar-politics/962ABE4FC41A6FF3E1F95CE1B54D1ADD)

```{r, echo = FALSE, warning = FALSE, message = FALSE}

# Read federal grants dataset from Kriner & Reeves (2008). 
d_pork_state <- read_csv("fedgrants_bystate_1988-2008.csv")

# What strategy do presidents pursue? 
d_pork_state |> 
  filter(!is.na(state_year_type)) |> 
  group_by(state_year_type) |>
  summarize(mean_grant = mean(grant_mil, na.rm = T), se_grant = sd(grant_mil, na.rm = T)/sqrt(n())) |> 
  ggplot(aes(x = state_year_type, y = mean_grant, ymin = mean_grant-1.96*se_grant, ymax = mean_grant+1.96*se_grant)) + 
  coord_flip() + 
  geom_bar(stat = "identity", fill = "darkgreen") + 
  geom_errorbar(width = 0.2) + 
  labs(x = "Type of State & Year", 
       y = "Federal Grant Spending (Millions of $)", 
       title = "Federal Grant Spending (Millions $) by State Election Type") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))

```

As is visible, swing states appear to have historically received more federal grant funding than non-swing states. While not statistically significant, it is plausible that, even within swing states, more federal funding is received during election years than non-election years.

To further test this **pork** theory, I will run a regression at the county level. I am interested in determining whether there is a statistically significant relationship between incumbent vote swing and federal grant spending and if this relationship differs for districts in competitive and non-competitive states. A state will be considered competitive if the losing candidate averaged at least 45% in the last three presidential elections. 

$$IncVoteSwing_t = \beta_0 + \beta_1(\Delta FedGrantSpend_t \text{ x }CompetitiveState_t) + \beta_2\Delta FedGrantSpend + \beta_3CompetitiveState_t +\beta_t \text{Year Fixed Effects} + \delta_t$$
I will also control for additional variables like Iraq war casualties, % change in county-level per capita income, and ad spending to mitigate omitted variable bias.

```{r}

# Pork county model. 
d_pork_county <- read_csv("fedgrants_bycounty_1988-2008.csv")

pork_mod_county_2 <- lm(dvoteswing_inc ~ dpct_grants*comp_state + as.factor(year) +
                          dpc_income + inc_ad_diff + inc_campaign_diff + 
                          dhousevote_inc + iraq_cas2004 + iraq_cas2008 + 
                          dpct_popl,
                        data = d_pork_county)

# I used Chat GPT for help here

stargazer(pork_mod_county_2, type = "text", 
          title = "County Model Regression Results",
          dep.var.labels = "Incumbent Vote Swing",
          covariate.labels = c("Federal Grants (%)", 
                               "Competitive State", 
                               "Federal Grants * Competitive State"),
          omit = c("year", "dpc_income", "inc_ad_diff", "inc_campaign_diff", 
                    "dhousevote_inc", "iraq_cas2004", "iraq_cas2008", "dpct_popl"),
          star.cutoffs = c(0.05, 0.01, 0.001), 
          notes = "Standard errors are in parentheses.", 
          omit.stat = c("f", "ser"))
```

After controlling for potential confounders, running the regression, and outputting the variables of interest, we can note the following:

1) An additional percentage point increase in federal grants translates to a small but statistically significant increase in incumbent vote swing on average.

2) Counties in competitive states tend to contribute 0.155 more percentage points to incumbent vote swing than those in non-competitive states.

3) The effect of federal grant spending on incumbent vote swing is greater in competitive states.

All of these main coefficients of interest are statistically significant, and we can conclude that federal grant, *pork*, spending, does have a significant effect on incumbent vote share, and that it is greater in swing states. This would help explain why *pork* seems to go disproportionately to swing states.

## Application

Pork visualized

Time for Change Model

interesting election where both candidates are incumbents...but also no

# Citations:


Hale, Edward Everett, et al. **The Man without a Country**. Little, Brown, and Company, 1898.

Kriner, Douglas L., and Andrew Reeves “Presidential Particularism and Divide-the-Dollar Politics.” *American Political Science Review* 109.1 (2015): 155–171. Web.

“Reelection Rates over the Years.” *OpenSecrets*, www.opensecrets.org/elections-overview/reelection-rates. Accessed 28 Sept. 2024. 

# Data Sources: 

Data are from the US presidential election popular vote results from 1948-2020, [Kriner and Reeves 2015](https://www-cambridge-org.ezp-prod1.hul.harvard.edu/core/journals/american-political-science-review/article/presidential-particularism-and-dividethedollar-politics/962ABE4FC41A6FF3E1F95CE1B54D1ADD), and [the Federal Reserve Bank of St. Louis](https://fred.stlouisfed.org/).